---
title: "Assignment 2 - Wrangling Danish Kings"
author: "Asta Hoegh Gregersen"
date: "2026-12-03"
output: html_document
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load the kings

1. Look at the dataset that are you loading and check what its columns are separated by? (hint: open it in plain text editor to see)
 
The dataset I am loading contains columns separated by the seperator ;  


2. Create a `kings` object in R with the different functions below and inspect the different outputs.

- read.csv()´
-  `read_csv()`
-  `read.csv2()` 
-  `read_csv2()`
```{r eval=TRUE}
# Reviewing the output of different functions 
# Loading the library 
library(tidyverse)

kings1 <- read.csv("data/kongerækken_helt_ny_13_3.csv")
kings1

kings2 <- read_csv("data/kongerækken_helt_ny_13_3.csv")
kings2

kings3 <- read.csv2("data/kongerækken_helt_ny_13_3.csv")
kings3

kings4 <- read_csv2( "data/kongerækken_helt_ny_13_3.csv")
kings4
```
Answer: 
1. Which of these functions is a `tidyverse` function? Read data with it below into a `kings` object
The functions read_csv and read_csv2 are tidyverse functions. When used, they create tibbles. As the data in my dataset is separeted by semicolons, I use the read_csv2 function to create my 'kings' object.  

2. What is the result of running `class()` on the `kings` object created with a tidyverse function.

One is able to the class of the object 'kings'. One gets the result spec_tnl_df "tbl_df" "tbl" and "data.frame". This is how, when data is read using the function read_csv2, the data is stored. 


3. How many columns does the object have when created with these different functions?

Using read.csv, the object has 1 column.
Using read_csv, the object once again has 1 column. 
Using read.csv2, the object has 7 columns. 
Using read_csv2, the object has 7 columns. 


4. Show the dataset so that we can see how R interprets each column

```{r eval = TRUE}
# My code 

kings <- read_csv2("data/kongerækken_helt_ny_13_3.csv")

class(kings)

# Showcasing the dataset
```
```{r eval = TRUE}
view(kings)


glimpse(kings)


```


## Calculate the duration of reign for all the kings in your table


```{r}
# My code 

kings <- kings %>%
  mutate(duration=last_year_ruled - first_year_ruled)

kings[[8]]


```

## Calculate the average duration of reign for all rulers


```{r}
# Using the mean() function on the duration column as a vector 
kings[[8]]
mean(kings$duration,na.rm=TRUE)
# One gets the average 19,42308


# Using the summarise() function on the duration column as a tibble

kings %>% 
  summarise(mean_duration=mean(duration,na.rm=TRUE))
# One gets a new tibble with 1 row, that row is called mean_duration, it contains the average reign, and it is 19,42308


```

## How many and which kings enjoyed a longer-than-average duration of reign?


```{r}
# My code
filter(kings,duration>19.42308)
# The names of these 24 kings can be viewed using the above code 

count(filter(kings,duration>19.42308))
# 24 kings enjoyed a longer-than-average duration of reign


```


## How many days did the three longest-ruling monarchs rule? 


```{r}
# My code 

arrange(kings,desc(duration))

arrange(kings,desc(duration)) %>% 
  slice(1:3)
# The three longest-ruling monarchs were Margrethe the 2nd, Christian the 4th and Christian the 9th. 


# Creating the column 'days'

kings <- kings %>% 
  mutate(days=duration*365)
view(kings)
         
# A column containing every kings reign in days has been added.
# One can now use the arrange() function, and then desc(), to get the amount of days each king has reigned in descending order. 

arrange(kings,desc(days)) %>% 
  slice(1:3)

# Margrethe the 2nd ruled for 18980 days, Christian the 4th reigned for 18980 days, and Christian the 9th reigned for 15695 days. 

```


## Plotting the kings' duration of reign through time

What is the long-term trend in the duration of reign among Danish monarchs? 

```{r}
kings %>% mutate(duration=last_year_ruled-first_year_ruled) %>%
  mutate(mid_year=(first_year_ruled+(last_year_ruled-first_year_ruled)/2))%>%
  ggplot(aes(x=mid_year,y=duration))+
  geom_point(color="red")+
               geom_smooth()+ 
  labs(title="Kings' duration of reign through time")
               theme_minimal()
```
Looking at the resulting plot, a clear trend emerges. Since the first kings, the duration of reign has steadily increased. After 1500, more kings reigned longer than the average duration of reigns, than they did before. 

