---
title: "Final Project: Absolutism in Denmark and Sweden"
author: "Asta Hoegh Gregersen"
date: "2026-05-16"
output: html_document
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
```

# Thoughts before starting the analysis
As has been outlined in the written part of this assignment, I believe that absolutism greatly benefited the monarchs of Denmark and Sweden, and led to longer reigns on average. Absolutism clearly defined the power of the monarchs, and put a stop to struggles over the throne. This meant that monarchs lived longer, and were thus able to reign longer. The effects of a long-lasting monarch and few battles over succession, lead to stability within the nation for a time. Some in society probably criticized the monarchy and its absolute powers, but I argue that absolutism did benefit society in many ways. Absolutism was in effect in Denmark for almost 200 years, so it is supremely relevant to focus on the Danish monarchy when exploring the effects of absolutism.  
Absolutism in Sweden is very interesting. Unlike many European nations, absolutism in Sweden only lasted for 47 years. Despite this, the reigns of the two Swedish absolutist monarchs were considerably longer than the average duration of reign for a Swedish monarch before absolutism. The unique circumstances under which absolutism was introduced in Sweden will be explored further in the written part of the assignment. 
As has been outlined, I strongly suspect that the average duration of reign for the monarchs who ruled during absolutism in both nations will be higher than the average for all monarchs, and significantly higher than the average for the monarchs that ruled before absolutism. 

## Loading relevant packages 
```{r}
library(tidyverse)
library(here)
```


## Creating relevant folders
```{r}
dir.create("data")
dir.create("figures")
```


## Creating dataframes from the datasets
```{r}
danish_monarchs <-read.csv2("data/danish_monarchs_dataset_2026.CSV")
danish_monarchs
```


```{r}
swedish_monarchs <-read.csv2("data/swedish_monarchs_dataset_2026.CSV")
swedish_monarchs
```


## Creating vectors of the columns of the Danish monarchs 
```{r}
regent_denmark<-danish_monarchs[[1]]
gender_denmark<-danish_monarchs[[2]]
birth_denmark<-danish_monarchs[[3]]
death_denmark<-danish_monarchs[[4]]
first_year_ruled_denmark<-danish_monarchs[[5]]
last_year_ruled_denmark<-danish_monarchs[[6]]
```


## Creating vectors of the columns of the Swedish monarchs 
```{r}
regent_sweden<-swedish_monarchs[[1]]
gender_sweden<-swedish_monarchs[[2]]
birth_sweden<-swedish_monarchs[[3]]
death_sweden<-swedish_monarchs[[4]]
first_year_ruled_sweden<-swedish_monarchs[[5]]
last_year_ruled_sweden<-swedish_monarchs[[6]]
```


## Creating a new column in each dataset 
This assignment has a focus on duration of reign in relation to absolutism. To examine this, a column containing the number of years each monarch reigned is required. Within both dataframes, a new column is created, containing the number of years each monarch ruled. The columns are going to be called no_years_ruled in each dataframe. A vector is then created of each of these columns, and the work with the data can begin. 

```{r}
danish_monarchs<-danish_monarchs%>%mutate(no_years_ruled=last_year_ruled-as.numeric(first_year_ruled))
         
no_years_ruled_denmark<-danish_monarchs[[7]]

no_years_ruled_denmark

```


```{r}
swedish_monarchs<-swedish_monarchs%>%mutate(no_years_ruled=last_year_ruled-as.numeric(first_year_ruled))
         
no_years_ruled_sweden<-swedish_monarchs[[7]]

no_years_ruled_sweden
```


# Working with the data
The hypothesis at the center of this project, is that monarchs, whether Danish or Swedish, on average reigned longer during absolutism. The average duration of reign overall, before and during absolutism will be found. Then, the different averages will be compared. 

# Average duration of reign in Denmark in relation to absolutism 

The column 'no_years_ruled' has been made into a vector, and then the average duration of reign across all monarchs can be found, using the mean() function. 
```{r}
mean(no_years_ruled_denmark,na.rm=TRUE)
```
Thus, the overall average duration of reigns for the Danish monarchs are 19,4 years. 

## Creating a dataframe containing only the Danish monarchs that reigned during absolutism

Now, two new data frames will be created. The first will contain the monarchs who ruled during absolutism. The second will contain the monarchs who ruled before absolutism. This will firstly be done with the Danish monarchs, and then with the Swedish monarchs in the next part of the assignment. 

After having conducted research about the Danish monarchs that reigned during absolutism, I know that the monarch who introduced absolutism was Frederik the 3rd, whose reign began in 1648. The last king to rule during absolutism was Christian the 8th. His death in 1848 marked the end of absolutism in Denmark. 

The filter() function is used to create the new data frame. 
Using first_year_ruled_denmark>1647, any monarch who reigned before 1647 is excluded, as Frederik the 3rd ascended the throne in 1648. 
Using last_year_ruled_denmark<1849, any monarchs who ruled after 1849 are excluded and Christian the 8th is included, as the last year of his reign was 1848

no_years_ruled_denmark!= "NA" ensures that any monarchs whose reign does not have either a start or an end, due to insufficient historical knowledge, are not included in the dataframe.  
 
```{r}
danish_monarchs_during_absolutism <- filter(danish_monarchs,first_year_ruled_denmark>1647,
       last_year_ruled_denmark<1849, no_years_ruled_denmark!="NA")
danish_monarchs_during_absolutism
```
Now that a new dataframe has been created, one can make a vector of the column no_years_ruled, and then use the mean() function to find the average duration of reign for the 8 monarchs that reigned during absolutism. 

```{r}
no_years_ruled_during_absolutism_denmark <- danish_monarchs_during_absolutism[[7]]

no_years_ruled_during_absolutism_denmark

mean(no_years_ruled_during_absolutism_denmark,na.rm=TRUE)
```
Thus, the average duration of reign during absolutism was 25, which is 5,6 years higher than the average duration of reign for all the Danish monarchs. 

Now, a third data frame will be created, which contains only the monarchs that reigned before absolutism. Then one is able to compare the overall average duration of reign, the average duration of reign during absolutism and the average duration of reign before absolutism. 

## Creating a dataframe containing only the Danish monarchs that reigned before absolutism 

The same functions as was used to create the previous data frame, is employed. first_year_ruled_denmark<1647 ensured that only the kings who reigned before Frederik the 3rd ascended the throne in 1648, are included. 
```{r}
danish_monarchs_before_absolutism <- filter(danish_monarchs,first_year_ruled_denmark<1647, no_years_ruled_denmark!="NA")

danish_monarchs_before_absolutism
```

Then, the column containing the number of years reigned can be made into a vector, and the mean() function can be used to find the average duration of reign for the Danish monarchs before absolutism. 
```{r}
no_years_ruled_before_absolutism_denmark <- danish_monarchs_before_absolutism[[7]]

no_years_ruled_before_absolutism_denmark

mean(no_years_ruled_before_absolutism_denmark,na.rm=TRUE)
```
The average duration of reign before absolutism was 16,7 years. This is 2,7 years shorter than the overall average, and 8,3 years shorter than the average duration of reign during absolutism. Thus, the monarchs that ruled during absolutism did indeed ruled longer on average than the monarchs who ruled before absolutism. They also ruled longer than the overall average across all Danish monarchs through time. 


# Average duration of reign in Sweden in relation to absolutism 
In order to compare absolutism in Denmark and Sweden, the same calculations must be performed. 

```{r}
mean(no_years_ruled_sweden,na.rm=TRUE)
```
Thus, the overall median duration of reigns for the Swedish monarchs are 13,1 years. 

## Creating a dataframe containing only the Swedish monarchs that reigned during absolutism

Now, as has been done previously, two new data frames will be created. The first will contain the monarchs who ruled during absolutism in Sweden. The second will contain the Swedish monarchs who ruled before absolutism.

After having conducted research about absolutism in Sweden, I know that absolutism oficially began in 1672, when Karl the 11th became of legal age, and ended in 1719, after Karl the 12th death in 1718 

The filter() function is once again used.
Using first_year_ruled_sweden>1659, any monarch who reigned before 1659 is excluded, as Karl the 11th ascended the throne in 1659. 
Using last_year_ruled_sweden<1720, any monarchs who ruled after 1720 are excluded and Karl the 12th is included, as his last year of reign was 1719.
 
```{r}
swedish_monarchs_during_absolutism <- filter(swedish_monarchs,first_year_ruled_sweden>1659,
       last_year_ruled_sweden<1720, no_years_ruled_sweden!="NA",birth!="NA")

swedish_monarchs_during_absolutism
```
Now that a new dataframe has been created, one can make a vector of the column duration_sweden, and then use the mean() function to find the average duration of reign for the 2 monarchs that reigned during absolutism in Sweden. 

```{r}
no_years_ruled_during_absolutism_sweden <- swedish_monarchs_during_absolutism[[7]]

no_years_ruled_during_absolutism_sweden

mean(no_years_ruled_during_absolutism_sweden,na.rm=TRUE)
```

Thus, the average duration of reign during absolutism in Sweden was 29, which is 15,9 years higher than the average duration of reign for all the Swedish monarchs. 

Now, a third data frame will be created, which contains only the monarchs that reigned before absolutism. Then the same results have been found across both nations, and then one can begin to compare and contrast absolutism in Denmark and Sweden. 


## Creating a dataframe containing only the Swedish monarchs that reigned before absolutism 

The same functions as was used to create the previous data frame, is employed. first_year_ruled_sweden<1659 ensures that only the monarchs who reigned before Karl the 11th are included. 
```{r}
swedish_monarchs_before_absolutism <- filter(swedish_monarchs,first_year_ruled_sweden<1659, no_years_ruled_sweden!="NA")

swedish_monarchs_before_absolutism
```


Then, the column containing the number of years reigned can be made into a vector, and the mean() function can be used to find the average duration of reign for the Swedish monarchs before absolutism. 
```{r}
no_years_ruled_before_absolutism_sweden <- swedish_monarchs_before_absolutism[[7]]

no_years_ruled_before_absolutism_sweden

mean(no_years_ruled_before_absolutism_sweden,na.rm=TRUE)
```
The average duration of reign before absolutism in Sweden was 11,3 years. This is 1,6 years shorter than the overall average, and a significant 17,7 years shorter than the average duration of reign during absolutism. 


## Comparing the results 
After having conducted the same calculations in both dataframes, it becomes clear that the average duration of reign during absolutism was much higher in both Denmark and Sweden. In Denmark, the average duration of reign during absolutism was 8,3 years higher than the average duration of reign before absolutism. In Sweden, the average duration of reign was a signifcant 17,7 years longer than the average duration of reign before absolutism. These calculations thus prove the hypothesis at the center of this assignment, that absolutism brought stability and longer reigns to their respective nations. 


# Plotting the results 
To support my hypothesis, and to properly showcase the effect of absolutism in the two nations, different plots will be created. Firstly, plots showcasing data relating Danish absolutism, and then plots showcasing data relating to Swedish absolutism.


# Visualizing the findings relating to absolutism in Denmark

## Duration of reign during absolutism, Denmark - duration over time
```{r}
denmark_absolutism_dot <- danish_monarchs_during_absolutism %>% mutate(no_years_ruled) %>%
  mutate(mid_year=(as.numeric(first_year_ruled)+(no_years_ruled)/2))%>%
  ggplot(aes(x=mid_year,y=no_years_ruled))+
  geom_point(color="purple")+
               geom_smooth()+ 
  labs(title="Danish monarchs' duration of reign during absolutism")+
  labs(y="Number of years ruled", x="Year")

# Showing the plot 
denmark_absolutism_dot

# Saving as a png 
ggsave(plot = denmark_absolutism_dot, 
       here("figures","denmark_absolutism_dot.png"), 
       height = 6, 
       width = 8)

```


## Duration of reign before absolutism, Denmark - duration over time
```{r}
denmark_before_absolutism_dot <- danish_monarchs_before_absolutism %>% mutate(no_years_ruled) %>%
  mutate(mid_year=(as.numeric(first_year_ruled)+(no_years_ruled)/2))%>%
  ggplot(aes(x=mid_year,y=no_years_ruled))+
  geom_point(color="purple")+
               geom_smooth()+ 
  labs(title="Duration of reign before absolutism, Denmark")+
  labs(x="Year",y="Number of years ruled")

# Showing the plot 
denmark_before_absolutism_dot

# Saving as a png 
ggsave(plot = denmark_before_absolutism_dot, 
       here("figures","denmark_before_absolutism_dot.png"), 
       height = 6, 
       width = 8)

```


## Average duration of reign before absolutism, Denmark - barplot 
```{r,fig.width=6,fig.height=8}
denmark_before_absolutism_barplot <-ggplot(danish_monarchs_before_absolutism, aes(x=regent,y=no_years_ruled))+geom_bar(stat="identity") + coord_flip()+
  labs(title="Duration of reign before absolutism, Denmark")+
  labs(x="Monarchs",y="Number of years ruled")

# Showing the plot 
denmark_before_absolutism_barplot

# Saving the plot as a png 
ggsave(plot = denmark_before_absolutism_barplot, 
       here("figures","denmark_before_absolutism_barplot.png"), 
       height = 8, 
       width = 6)
```


## Average duration of reign during absolutism, Denmark - barplot 
```{r}
denmark_absolutism_barplot <- ggplot(danish_monarchs_during_absolutism, aes(x=regent,y=no_years_ruled))+geom_col(aes(fill=regent))+
  labs(title="Duration of reign during absolutism, Denmark")+
  labs(x="Monarchs",y="Number of years ruled")

# Showing the plot 
denmark_absolutism_barplot

# Saving as a png 
ggsave(plot = denmark_absolutism_barplot, 
       here("figures","denmark_absolutism_barplot.png"), 
       height = 6, 
       width = 8)

```


# Visualizing the findings relating to absolutism in Sweden 

## Average duration of reign in Sweden before absolutism - duration over time
```{r}
sweden_before_absolutism_dot <- swedish_monarchs_before_absolutism %>% mutate(no_years_ruled) %>%
  mutate(mid_year=(as.numeric(first_year_ruled)+(no_years_ruled)/2))%>%
  ggplot(aes(x=mid_year,y=no_years_ruled))+
  geom_point(color="purple")+
               geom_smooth()+ 
  labs(title="Duration of reign before absolutism, Sweden")+
  labs(x="Year",y="Number of years ruled")

# Showing the plot 
sweden_before_absolutism_dot

# Saving as a png 
ggsave(plot = sweden_before_absolutism_dot, 
       here("figures","sweden_before_absolutism_dot.png"), 
       height = 6, 
       width = 8)

```


## Duration of reign in Sweden before absolutism - barplot 
```{r,fig.width=6,fig.height=8}
sweden_before_absolutism_barplot <- ggplot(swedish_monarchs_before_absolutism, aes(x=regent,y=no_years_ruled))+geom_bar(stat="identity")+ coord_flip()+labs(title="Duration of reign before absolutism, Sweden")+
  labs(x="Monarchs",y="Number of years ruled")

# Showing the plot 
sweden_before_absolutism_barplot

# Saving as a png 
ggsave(plot = sweden_before_absolutism_barplot, 
       here("figures","sweden_before_absolutism_barplot.png"), 
       height = 8, 
       width = 6)

```


## Duration of reign in Sweden during absolutism - barplot
```{r}
sweden_absolutism_barplot <- ggplot(swedish_monarchs_during_absolutism, aes(x=regent,y=no_years_ruled))+geom_col(aes(fill=regent))+labs(title="Duration of reign during absolutism, Sweden")+
  labs(x="Monarchs",y="Number of years ruled")

# Showing the plot 
sweden_absolutism_barplot

# Saving as a png 
ggsave(plot = sweden_absolutism_barplot, 
       here("figures","sweden_absolutism_barplot.png"), 
       height = 6, 
       width = 8)
```

# Conclusion

Thus, in Denmark the average duration of reign for all monarchs is 19,4 years, the average duration of reign for the monarchs that reigned during absolutism is 25 years and the average duration of reign before absolutism is 16,7 years. Looking at the monarchs who reigned before absolutism, and the ones that reigned during, it becomes clear that a much larger share of the monarchs who reigned during absolutism reigned much longer than the overall average. The same was not the case before absolutism, where an overwhelming amount of the monarchs reigned shorter than the overall average. Further commentary on the monarchs who ruled before and during absolutism can be found in the written part of the assignment. 
The same is true of Sweden, even though absolutism only spanned two Swedish monarchs. Both monarchs reigned many years longer than the average before absolutism, 11,3 years, something which was a rare feat before absolutism. 
The hypothesis at the center of this analysis, that the monarchs that reigned during absolutism reigned longer than the monarchs that reigned before absolutism, has been proved. 
An exploration of how absolutist was introduced in each nation will be found in the written assignment. Also, there will be an exploration of the plots created here, and the ways in which they support the central hypothesis. 