---
title: "Final_Project_Absolutism_in_Denmark"
author: "Asta Hoegh Gregersen"
date: "2026-05-16"
output: html_document
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
```

## Thoughts before starting the analysis
As has been outlined in the written part of this assignment, I strongly suspect that the average duration of reign for the monarchs who ruled during absolutism will be higher than the average for all monarchs, and higher than the average for the monarchs that ruled before absolutism. . 

# Loading relevant packages 
```{r packages}
library(tidyverse)
library(here)
```

# Creating relevant folders
```{r folder}
dir.create("data")
```

# Creating a dataframe 
```{r original dataframe}
danish_monarchs <-read.csv2("data/danish_monarchs_dataset.CSV")
danish_monarchs
```

# Creating vectors of the columns 
```{r vectors}
regent<-danish_monarchs[[1]]
gender<-danish_monarchs[[2]]
birth<-danish_monarchs[[3]]
death<-danish_monarchs[[4]]
first_year_ruled<-danish_monarchs[[5]]
last_year_ruled<-danish_monarchs[[6]]
no_years_ruled <- danish_monarchs[[7]]
```

The hypothesis at the center of this project, is that Danish monarchs reigned longer during absolutism, than before absolutism, and longer overall. To examine this claim, the column 'no_years_ruled' has been made into a vector, and then one can find the median duration of reign across all monarchs, using the mean() function. 
```{r overall average}
mean(no_years_ruled,na.rm=TRUE)
```
Thus, the overall median duration of reigns for the Danish monarchs are 19,4 years. 

# Creating a dataframe containing only the monarchs that reigned during absolutism

Now, two new data frames will be created. The first will contain the monarchs who ruled during absolutism. The second will contain the monarchs who ruled before absolutism.

After having done research about the Danish monarchs that reigned during absolutism, I know that the monarch who introduced absolutism was Frederik the 3rd, whose reign began in 1648, and the last king to rule during absolutism was Christian the 8th. His death in 1848 marked the end of absolutism in Denmark. 
The filter() function is used to create the new data frame. 
Using first_year_ruled>1647, any monarch who reigned before 1647is excluded, as Frederik the 3rd ascended the throne in 1648. 
Using last_year_ruled<1849, any monarchs who ruled after 1849 are excluded and Christian the 8th is included, as his last year of reign was 1848

no_years_ruled != "NA" ensures that any monarchs whose reign does not have either a first year or a last year, due to insufficient historical knowledge, are not included in the dataframe.  
 
```{r first new dataframe}
danish_monarchs_during_absolutism <- filter(danish_monarchs,first_year_ruled>1647,
       last_year_ruled<1849, no_years_ruled!="NA")
danish_monarchs_during_absolutism
```
Now that a new dataframe has been created, one can make a vector of the column no_years_ruled, and then use the mean() function to find the average duration of reign for the 8 monarchs that reigned during absolutism. 

```{r absolutism average}
no_years_ruled_during_absolutism <- danish_monarchs_during_absolutism[[7]]
no_years_ruled_during_absolutism
mean(no_years_ruled_during_absolutism,na.rm=TRUE)
```
Thus, the average duration of reign during absolutism was 25, which is 5,6 years higher than the average duration of reign for all the Danish monarchs. 

Now, a third data frame will be created, which contains only the monarchs that reigned before absolutism. Then one is able to compare the overall average duration of reign, the average duration of reign during absolutism and the average duration of reign before absolutism. 

# Creating a dataframe containing only the monarchs that reigned before absolutism 

The same functions as was used to create the previous data frame, is employed. first_year_ruled<1647 ensured that only the kings who reigned before Frederik the 3rd ascended the throne in 1948, are included. 
```{r second new dataframe}
danish_monarchs_before_absolutism <- filter(danish_monarchs,first_year_ruled<1647, no_years_ruled!="NA")
danish_monarchs_before_absolutism
```

Then, the column containing the number of years reigned can be made into a vector, and the mean() function can be used to find the average duration of reign for the Danish monarchs before absolutism. 
```{r before absolutism average}
no_years_ruled_before_absolutism <- danish_monarchs_before_absolutism[[7]]
no_years_ruled_before_absolutism
mean(no_years_ruled_before_absolutism,na.rm=TRUE)
```
The average duration of reign before absolutism was 16,7 years. 

# Longest reigning monarch during absolutism 

```{r}
max(no_years_ruled_during_absolutism,na.rm=TRUE)
```
The monarch who, during absolutism, reigned for 42 years, is Christian the 7th. A closer examination of the unique reign of Christian the 7th can be found in the portfolio. 

# Longest reigning monarch before absolutism 
```{r}
max(no_years_ruled_before_absolutism,na.rm=TRUE)
```
The monarch who reigned for 52 years is Christian the 4th. He is the longest reigning monarch before absolutism. An exploration of his reign, and how absolutism possibly could have changed it for the better, can be found in the portfolio. 

## Conclusion

Thus, the average duration of reign for all monarchs is 19,4 years, the average duration of reign for the monarchs that reigned during absolutism is 25 years and the average duration of reign before absolutism is 16,7 years. The hypothesis at the center of this analysis, that the monarchs that reigned during absolutism reigned longer than the monarchs that reigned before absolutism, has been proved. The monarchs that reigned during absolutism ruled 5,6 years longer than the overall average, and 8,3 years longer than the average of the monarchs that reigned before absolutism. 

This ways in which absolutism promoted longer and less troubled reigns, will be explored in the written part of the assignment. There will also be an examination of the kings highlighted in this analysis: Christian the 7th, Knud the 2nd, Svend Grathe, Valdemar the 1st and lastly, Christian the 4th. 
